---
title: "Getting Started with CellDeathAnalysis"
author: "Keran Sun"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CellDeathAnalysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

**CellDeathAnalysis** is an R package designed for comprehensive analysis of 
cell death pathway-related genes in transcriptomic data. This vignette provides 
a quick introduction to the main features of the package.

### Supported Cell Death Types

The package includes curated gene sets for 14 types of regulated cell death:

1. **Ferroptosis** (铁死亡) - Iron-dependent cell death
2. **Cuproptosis** (铜死亡) - Copper-dependent cell death  
3. **Disulfidptosis** (二硫死亡) - Disulfide stress-induced death
4. **Pyroptosis** (焦亡) - Inflammatory cell death
5. **Necroptosis** (坏死性凋亡) - Programmed necrosis
6. **Apoptosis** (凋亡) - Classical programmed cell death
7. **Autophagy** (自噬) - Self-degradation process
8. **PANoptosis** (泛凋亡) - Combined cell death
9. **NETosis** - Neutrophil extracellular trap-associated death
10. **Parthanatos** - PARP-dependent cell death
11. **Entosis** - Cell-in-cell death
12. **Oxeiptosis** - Oxidation-induced death
13. **Alkaliptosis** - Alkaline pH-induced death
14. **LDCD** - Lysosome-dependent cell death

## Installation

```{r install, eval=FALSE}
# From GitHub
devtools::install_github("yourusername/CellDeathAnalysis")

# Load the package
library(CellDeathAnalysis)
```

```{r load, echo=FALSE, message=FALSE}
# For vignette building
# library(CellDeathAnalysis)
```

## Exploring Gene Sets

### List Available Pathways

```{r list_pathways, eval=FALSE}
# Simple list
list_death_pathways()

# Detailed information
info <- list_death_pathways(detailed = TRUE)
head(info)
```

### Get Pathway Information

```{r pathway_info, eval=FALSE}
# Get detailed information about a pathway
get_pathway_info("ferroptosis")
```

### Retrieve Gene Sets

```{r get_genes, eval=FALSE}
# Get full gene set (with categories)
ferro_full <- get_death_geneset("ferroptosis", type = "full")
names(ferro_full)

# Get simplified core genes
ferro_simple <- get_death_geneset("ferroptosis", type = "simple")
head(ferro_simple)

# Get all genes as a vector
ferro_all <- get_death_geneset("ferroptosis", type = "all")
length(ferro_all)

# Get specific category
ferro_drivers <- get_death_geneset("ferroptosis", category = "driver")
```

### Multiple Pathways

```{r multiple_pathways, eval=FALSE}
# Get multiple pathways at once
genesets <- get_death_geneset(
  c("ferroptosis", "pyroptosis", "cuproptosis"),
  type = "simple"
)

# Check overlap between pathways
overlap <- check_pathway_overlap(c("ferroptosis", "apoptosis", "autophagy"))
```

## Calculating Pathway Scores

### Prepare Expression Data

Your expression data should be:
- A matrix or data frame
- Genes as rows, samples as columns
- Row names should be gene symbols (HGNC format)

```{r prepare_data, eval=FALSE}
# Example: Create mock expression data
set.seed(123)
expr <- matrix(rnorm(500 * 100), nrow = 500, ncol = 100)
rownames(expr) <- sample(get_all_death_genes(), 500)
colnames(expr) <- paste0("Sample_", 1:100)
```

### Calculate Scores

```{r calculate_scores, eval=FALSE}
# Method 1: Z-score (fast, no dependencies)
scores_zscore <- calculate_death_score(
  expr, 
  pathways = "all",
  method = "zscore"
)

# Method 2: ssGSEA (requires GSVA package)
scores_ssgsea <- calculate_death_score(
  expr,
  pathways = c("ferroptosis", "pyroptosis"),
  method = "ssgsea"
)

# Method 3: Mean expression
scores_mean <- calculate_death_score(
  expr,
  method = "mean"
)
```

### View Results

```{r view_scores, eval=FALSE}
# Print summary
print(scores_zscore)

# View first few rows
head(scores_zscore)

# Summary statistics
summary(scores_zscore)
```

## Visualization

### Heatmap

```{r heatmap, eval=FALSE}
# Basic heatmap
plot_death_heatmap(scores_zscore)

# With sample annotations
sample_info <- data.frame(
  Group = rep(c("Control", "Treatment"), each = 50),
  Stage = sample(c("I", "II", "III", "IV"), 100, replace = TRUE)
)
plot_death_heatmap(scores_zscore, annotation_col = sample_info)
```

### Boxplot Comparison

```{r boxplot, eval=FALSE}
# Compare groups
group <- rep(c("Control", "Treatment"), each = 50)
plot_death_boxplot(scores_zscore, group = group)

# Select specific pathways
plot_death_boxplot(
  scores_zscore, 
  group = group,
  pathways = c("ferroptosis", "pyroptosis", "cuproptosis"),
  ncol = 3
)
```

### Radar Chart

```{r radar, eval=FALSE}
# Radar chart comparing groups
plot_death_radar(scores_zscore, group = group)
```

### Correlation Plot

```{r correlation, eval=FALSE}
# Correlation between pathways
plot_pathway_correlation(scores_zscore, method = "spearman")
```

### Score Distribution

```{r distribution, eval=FALSE}
# Density plot
plot_score_distribution(scores_zscore, type = "density")

# Violin plot by group
plot_score_distribution(scores_zscore, type = "violin", group = group)
```

## Sample Classification

### Classify by Score

```{r classify, eval=FALSE}
# Classify by median
ferro_groups <- classify_by_score(scores_zscore$ferroptosis, method = "median")
table(ferro_groups)

# Classify by quantile
ferro_groups_q <- classify_by_score(
  scores_zscore$ferroptosis, 
  method = "quantile",
  quantile = 0.75
)
```

### Calculate Death Index

```{r death_index, eval=FALSE}
# PCA-based composite index
death_index <- calculate_death_index(scores_zscore, method = "pca")

# Weighted mean
weights <- c(ferroptosis = 2, pyroptosis = 1, apoptosis = 1)
death_index_weighted <- calculate_death_index(
  scores_zscore[, names(weights)],
  weights = weights,
  method = "weighted_mean"
)
```

## Integration with Other Tools

### Survival Analysis

```{r survival, eval=FALSE}
# Requires survival and survminer packages
library(survival)
library(survminer)

# Prepare data
surv_data <- data.frame(
  time = clinical_data$survival_time,
  status = clinical_data$status,
  ferro_group = classify_by_score(scores_zscore$ferroptosis)
)

# Kaplan-Meier analysis
fit <- survfit(Surv(time, status) ~ ferro_group, data = surv_data)
ggsurvplot(fit, data = surv_data, pval = TRUE)
```

### Enrichment Analysis

```{r enrichment, eval=FALSE}
# Get genes for enrichment
ferro_genes <- get_death_geneset("ferroptosis", type = "all")

# Use with clusterProfiler
library(clusterProfiler)
library(org.Hs.eg.db)

# Convert to Entrez IDs and perform GO enrichment
# ... (your analysis code)
```

## Best Practices

1. **Data Preprocessing**: Ensure your expression data is properly normalized
   (TPM, FPKM, or log-transformed counts).

2. **Gene Symbol Format**: Use official HGNC gene symbols for best matching.

3. **Minimum Gene Overlap**: Pathways with fewer than 5 detected genes are 
   automatically excluded.

4. **Method Selection**:
   - Use `ssgsea` or `gsva` for publication-quality results
   - Use `zscore` or `mean` for quick exploratory analysis

5. **Multiple Testing**: When comparing many pathways, consider adjusting 
   p-values for multiple testing.

## Session Info

```{r session, eval=FALSE}
sessionInfo()
```

## References

1. Dixon SJ, et al. Ferroptosis: An Iron-Dependent Form of Nonapoptotic 
   Cell Death. Cell. 2012;149(5):1060-1072.

2. Tsvetkov P, et al. Copper induces cell death by targeting lipoylated 
   TCA cycle proteins. Science. 2022;375(6586):1254-1261.

3. Liu X, et al. Actin cytoskeleton vulnerability to disulfide stress 
   mediates disulfidptosis. Nat Cell Biol. 2023;25(3):404-414.

4. Hänzelmann S, Castelo R, Guinney J. GSVA: gene set variation analysis 
   for microarray and RNA-Seq data. BMC Bioinformatics. 2013;14:7.
